---
title: "R data structure with stringr, forcats, lubridate, and tidyr"
subtitle: "PSYC 259: Principles of Data Science"
author: "John Franchak"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
---
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view","broadcast"))
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
# Set up custom theme
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  base_font_size = "20px",
  text_font_size = "1.5rem",
  header_h1_font_size = "3rem",
  header_h2_font_size = "2.5rem",
  header_h3_font_size = "2rem",
)
```

```{r, include = F}
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(icons) #https://fontawesome.com/v5.15/icons?d=gallery&p=2&m=free
```

# Today's tutorial

- Working with different data types
  * Vectors and lists
  * Strings
  * Factors
  * Dates
- Tidy data
  * Pivoting from/to wide/long format
  * Binding and joining data
 
.content-box-blue[
`r icons::fontawesome("github")` Follow along from the [Github repo](https://github.com/psych-259-data-science-2022/259-data-structure)
]

.footnote[.small[.bold[Last updated: `r Sys.Date()`]]]

---
# Common data types in R
- Numeric
  * integer: 1, 2, 3
  * double: 1.12124
  
- .brand-red[Character: “hello”]

- Logical: T/F (TRUE/FALSE)

- .brand-red[Datetime]

- .brand-red[Factor]

- .brand-red[Vectors and lists]

---
# What packages do we need?

```{r, warning=TRUE, message=TRUE}
library(tidyverse)
library(lubridate) #install if needed
```

---
# What packages do we need?

- Default tidyverse
  * `ggplot2`: graphing
  * `tibble`: use tibble-style dataframes
  * `tidyr`: shaping tibbles and columns
  * `purrr`: tools for automation
  * `dplyr`: selecting, filtering, and transforming dataframes
  * `stringr`: functions for manipulating strings
  * `forcats`: functions for manipulating factors
- Extras
  * `lubridate`: functions for manipulating dates

---
# Vectors and lists: Collections of values
- Vectors are type-homogeneous
  * Numeric vectors, `c(1,2,3,4,5)`, `1:5`
  * Character vectors,  `c("Hello", "Goodbye")`
  * Logical vectors, `c(T, T, T, F, T, F)`
  * Vector items can be named `c(x = 1, y = 2)`
- Lists are type-heterogeneous
  * Mix types, `list("1", 1, TRUE, 1:5)`
  * Tibbles are lists of vectors/lists of the same length

---
`r chunk_reveal("vectors", break_type = "auto", widths = c(2, 1), title = "# Vectors")`

```{r vectors, include = FALSE}
v <- c(1, 2, 3, 4, 5)
v
typeof(v)
length(v)
v[1] # Use brackets to access a vector element
v[3]
v <- c(v, 6, 7, 8) # use c to add items
v <- c(1, 2, 3, "4", "5", "6") 
v # You cannot mix types in a vector
v <- c(x = 1, y = 2, z = 3) # Named vector
v
v['y']

files <- list.files(path = "data_raw")
files[1] # Character vector
files[2]
length(files) 
```

---
`r chunk_reveal("lists", break_type = "auto", widths = c(2, 1), title = "# Lists")`

```{r lists, include = FALSE}
lt <- list(x = 1, y = "3", z = 1:5)
lt
lt$z # Access named list elements with $
lt[[1]] # Use double brackets to get list elements
lt[[3]] <- 5:7

# List of lists
lt <- list(lt1 = lt, lt2 = lt, lt3 = lt)
lt$lt1

model <- lm(height ~ mass, data = starwars)
typeof(model)
model$coefficients
```

---
# Factors represent categories
```{r}
x <- c(1, 1, 2, 3, 1, 4)
x <- factor(x, levels = c(1,2,3), labels = c("rarely", "neutral", "frequently"))
x
```
- levels restrict the possible set of values

- levels are ordered, which carries forward to output, modeling, graphics, etc.

- factors work as dummy codes; use as.numeric(factor) to treat as a continuous variable in models

---
`r chunk_reveal("factors", break_type = "auto", widths = c(1, 1), title = "# Factors")`

```{r factors, include = FALSE}
ds <- starwars
unique(ds$eye_color)
ds <- ds %>% mutate(eye_f = factor(eye_color))
fct_count(ds$eye_f, sort = T) %>% slice_head(n = 5)
fct_count(ds$eye_f, prop = T) %>% slice_head(n = 5)
```

---
`r chunk_reveal("releveling", break_type = "rotate", float = "none", widths = c(8, 1), title = "# Setting factor levels")`

```{r releveling, include = FALSE}
ds <- ds %>% select(name, eye_f)
ds #ROTATE
ds %>% mutate(eye_f2 = factor(eye_f, levels = c("blue", "brown"))) #ROTATE
ds %>% mutate(eye_f2 = factor(eye_f, levels = c("blue", "brown"), labels = c("bl","br"))) #ROTATE
ds %>% mutate(eye_f2 = fct_lump(eye_f, n = 3)) #ROTATE
ds %>% mutate(eye_type = fct_collapse(eye_f, human = c("blue", "brown"), nonhuman = c("yellow", "red"))) #ROTATE
```

<!-- adjust font size and other formatting defs in this css code chunk -->
```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1; font-size: 70%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}

.remark-slide-number {
  opacity: 0; /* default: 0.5 */
}

.content-box { 
    box-sizing: border-box;
    background-color: #e2e2e2;
}
.content-box-blue,
.content-box-gray,
.content-box-grey,
.content-box-army,
.content-box-green,
.content-box-purple,
.content-box-red,
.content-box-yellow {
  box-sizing: border-box;
  border-radius: 15px;
  margin: 0 0 15px;
  overflow: hidden;
  padding: 0px 20px 0px 20px;
  width: 100%;
}
.content-box-blue { background-color: #F0F8FF; }
.content-box-gray { background-color: #e2e2e2; }
.content-box-grey {	background-color: #F5F5F5; }
.content-box-army {	background-color: #737a36; }
.content-box-green { background-color: #d9edc2; }
.content-box-purple { background-color: #e2e2f9; }
.content-box-red { background-color: #ffcccc; }
.content-box-yellow { background-color: #fef5c4; }
.content-box-blue .remark-inline-code,
.content-box-blue .remark-inline-code,
.content-box-gray .remark-inline-code,
.content-box-grey .remark-inline-code,
.content-box-army .remark-inline-code,
.content-box-green .remark-inline-code,
.content-box-purple .remark-inline-code,
.content-box-red .remark-inline-code,
.content-box-yellow .remark-inline-code { 
  background: none;
}

.scroll-box-8 {
  height:8em;
  overflow-y: scroll;
}
.scroll-box-10 {
  height:10em;
  overflow-y: scroll;
}
.scroll-box-12 {
  height:12em;
  overflow-y: scroll;
}
.scroll-box-14 {
  height:14em;
  overflow-y: scroll;
}
.scroll-box-16 {
  height:16em;
  overflow-y: scroll;
}
.scroll-box-18 {
  height:18em;
  overflow-y: scroll;
}
.scroll-box-20 {
  height:20em;
  overflow-y: scroll;
}
.scroll-output {
  height: 90%;
  overflow-y: scroll;
}
}

/************************
 * Font size and colours
 ************************/

/*      LaTeX style       */
.Large       , .Large .remark-code, .Large .remark-inline-code { font-size: 144%; }
.large       , .large .remark-code, .large .remark-inline-code { font-size: 120%; }
.small       , .small .remark-code, .small .remark-inline-code { font-size: 90%; }
.footnotesize, .footnotesize .remark-code, .footnotesize .remark-inline-code { font-size: 80%; }
.scriptsize  , .scriptsize .remark-code, .scriptsize .remark-inline-code { font-size: 70%; }
.tiny        , .tiny .remark-code, .tiny .remark-inline-code { font-size: 60%; }

/* or you can be more specific */
.font10 , .code10 .remark-code, .code10 .remark-inline-code{ font-size: 10%; }
.font20 , .code20 .remark-code, .code20 .remark-inline-code{ font-size: 20%; }
.font30 , .code30 .remark-code, .code30 .remark-inline-code{ font-size: 30%; }
.font40 , .code40 .remark-code, .code40 .remark-inline-code{ font-size: 40%; }
.font50 , .code50 .remark-code, .code50 .remark-inline-code{ font-size: 50%; }
.font60 , .code60 .remark-code, .code60 .remark-inline-code{ font-size: 60%; }
.font70 , .code70 .remark-code, .code70 .remark-inline-code{ font-size: 70%; }
.font75 , .code75 .remark-code, .code75 .remark-inline-code{ font-size: 75%; }
.font80 , .code80 .remark-code, .code80 .remark-inline-code{ font-size: 80%; }
.font90 , .code90 .remark-code, .code90 .remark-inline-code{ font-size: 90%; }
.font100, .code100 .remark-code, .code100 .remark-inline-code{ font-size: 100%; }
.font110, .code110 .remark-code, .code110 .remark-inline-code{ font-size: 110%; }
.font120, .code120 .remark-code, .code120 .remark-inline-code{ font-size: 120%; }
.font130, .code130 .remark-code, .code130 .remark-inline-code{ font-size: 130%; }
.font140, .code140 .remark-code, .code140 .remark-inline-code{ font-size: 140%; }
.font150, .code150 .remark-code, .code150 .remark-inline-code{ font-size: 150%; }
.font160, .code160 .remark-code, .code160 .remark-inline-code{ font-size: 160%; }
.font170, .code170 .remark-code, .code170 .remark-inline-code{ font-size: 170%; }
.font175, .code175 .remark-code, .code175 .remark-inline-code{ font-size: 175%; }
.font180, .code180 .remark-code, .code180 .remark-inline-code{ font-size: 180%; }
.font190, .code190 .remark-code, .code190 .remark-inline-code{ font-size: 190%; }
.font200, .code200 .remark-code, .code200 .remark-inline-code{ font-size: 200%; }

.brand-red { color: #e64626; }
.brand-blue { color: #0148A4; }
.brand-yellow { color: #FFB800; }
.brand-charcoal {color: #424242; }
.brand-gray {color: #F1F1F1;}
.brand-grey {color: #F1F1F1;}
.black { color: black; }
.white { color: white; }
.red { color: red; }
.blue { color: blue; }
.green { color: green; }
.yellow { color: yellow; }
.orange { color: orange; }
.purple { color: purple; }
.gray { color: gray; }
.grey { color: gray; }

.bold { font-weight: bold; }
.bolder { font-weight: bolder; }
```

